---
title: "Stat 145 -- PS07"
author: "Thomas Scofield"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  pdf_document:
    fig_height: 2.2
    fig_width: 4
  html_document:
    fig_height: 2.2
    fig_width: 4
  word_document:
    fig_height: 2.2
    fig_width: 4
---

```{r, setup, include = FALSE, message=FALSE}
# load packages that are going to be used
library(mosaic)      # this loads ggformula (for plotting), etc. too
library(openintro)   # this loads data sets intro to modern stats 
library(oibiostat)   # this loads data sets open intro biostats 
library(pander)      # for tables

# Some customization. You can alter or delete as desired (if you know what you are doing).

theme_set(theme_bw())     # change theme for ggplot2/ggformula

knitr::opts_chunk$set(
  tidy = FALSE,     # display code as typed (rather than reformatted)
  size = "small")   # slightly smaller font for code
```

<!-- Put your work below here.  Put text in text chunks, code in R chunks. -->

### Exercise 2.100

How to make a list of values, name it *times*, and calculate its mean.
```{r}
times <- c(40, 87, 78, 106, 67, 70, 153, 81)
mean(~times)
```

### More examples of text, R code and results

To include a scatter plot of data in *faithful*, with *eruptions*
as explanatory variable and *waiting* as response variable.
```{r}
gf_point(waiting ~ eruptions, data=faithful)
```

This exemplifies a **positive association**, since the larger the value of the explanatory variable (*eruptions*), the larger the response tends to be.

A **negative association** does the opposite---the larger the explanatory value, the *smaller* the response tends to be.

## Correlation

Correlation measures the strength of a linear association.  For this data, the value of the correlation is
```{r}
cor(waiting ~ eruptions, data=faithful)
```

Some facts about correlation:

 * It requires two quantitative variables.
 * It's always a number between (-1) and (+1).
 * When close to zero, the *linear* association is weak.
 * When close to (-1) or to (+1) the linear association is strong.
 * It's positive or negative to match the slope of a regression line.
 * If you change the unit of measurement on one of the variables, this has no effect on the correlation.
 * It doesn't matter which variable is considered explanatory, and which is considered response.  Trade roles between the two variables, and the result is the same:
```{r}
cor(eruptions ~ waiting, data=faithful)
```
 