---
title: "RStudio Cheat Sheet"
author: "T.Scofield"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  pdf_document:
    fig_height: 2.2
    fig_width: 4
  html_document:
    fig_height: 2.2
    fig_width: 4
  word_document:
    fig_height: 2.2
    fig_width: 4
---

```{r, setup, include = FALSE, message=FALSE}
# load packages that are going to be used
library(mosaic)      # this loads ggformula (for plotting), etc. too
library(openintro)   # this loads data sets intro to modern stats 
library(oibiostat)   # this loads data sets open intro biostats 
library(pander)      # for tables

# Some customization. You can alter or delete as desired (if you know what you are doing).

theme_set(theme_bw())     # change theme for ggplot2/ggformula

knitr::opts_chunk$set(
  tidy = FALSE,     # display code as typed (rather than reformatted)
  size = "small")   # slightly smaller font for code
```

## Making a list of numbers, calculating statistics from it

Suppose I want things like the mean, sd, median, etc. for the
list of numbers 9, 11, 7, 13, 10.
```{r}
c(9, 11, 7, 13, 10) -> x
mean(~x)
```

## Tables
To make a frequency table
```{r}
ssurv <- read.csv("http://scofield.site/teaching/data/csv/ssurv.csv")
tally(~ selfhandedness, data=ssurv)
```

If I want a relative frequency table instead
```{r}
tally(~ selfhandedness, data=ssurv, format="proportion")
```

For bivariate data (i.e., two-way tables)
```{r}
tally(~ selfhandedness | sex, data=ssurv)
```

Adding the "format" switch
```{r}
tally(~ selfhandedness | sex, data=ssurv, format="percent")
```
gives us percentages out of the whole, done for each column.  If we want to reverse the roles of `sex` and `handedness`,
```{r}
tally(~ sex | selfhandedness, data=ssurv, format="percent")
```

Now, to make the total of all combined cells make up 100 percent,
```{r}
tally(~selfhandedness | sex, data=ssurv) %>% prop.table()
```

## Plotting bivariate data
In the `iris` data frame, there is a column (categorical) called
`Species` and another column (quantitative) called `Sepal.Width`.
If I want side-by-side boxplots for the quantitative variable
broken down by the categorical one
```{r}
gf_boxplot(~ Sepal.Width | Species, data=iris)
```
Now, try out this modification

```{r}
gf_boxplot(Species ~ Sepal.Width, data=iris)
```

