---
title: "Two-Proportion inference"
author: "Thomas Scofield"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  pdf_document:
    fig_height: 2.2
    fig_width: 4
  html_document:
    fig_height: 2.2
    fig_width: 4
  word_document:
    fig_height: 2.2
    fig_width: 4
---

```{r, setup, include = FALSE, message=FALSE}
# load packages that are going to be used
library(mosaic)      # this loads ggformula (for plotting), etc. too
library(openintro)   # this loads data sets intro to modern stats 
library(oibiostat)   # this loads data sets open intro biostats 
library(pander)      # for tables

# Some customization. You can alter or delete as desired (if you know what you are doing).

theme_set(theme_bw())     # change theme for ggplot2/ggformula

knitr::opts_chunk$set(
  tidy = FALSE,     # display code as typed (rather than reformatted)
  size = "small")   # slightly smaller font for code
```

First, a fact:

**Theorem**: Suppose $X$ and $Y$ are independent variables, and both
are normally distributed, with $X\sim \mbox{Norm}(\mu_X, \sigma_X)$
and $Y\sim \mbox{Norm}(\mu_Y, \sigma_Y)$.  Then their difference
$X-Y$ also has a normal distribution, with $(X-Y) \sim \mbox{Norm}(\mu_X
- \mu_Y, \sqrt{\sigma_X^2 + \sigma_Y^2})$.

### Two-Proportion context
Imagine you have two groups/populations in mind, and you take
*independent* samples, one of size $n_1$ from Group 1, and one
of size $n_2$ from Group 2.  The variable you measure is binary
categorical (sex, Christian or not?, have a certain gene or not?).
The proportions of *successes* are

 * $p_1$, $p_2$, in the two populations
 * $\widehat p_1$, $\widehat p_2$, in the two samples

Note that

 * $\widehat p_1$, $\widehat p_2$ should be independent, since
   the samples are.
 * If the rules-of-thumb
   $$ n_1 p_1 \ge 10 \qquad\mbox{and}\qquad n_1(1-p_1)
   \ge 10$$
   are met, then
   $$ p_1 \sim \mbox{Norm}\left(p_1,
   \sqrt{\displaystyle{\frac{p_1(1-p_1)}{n_1}}}\right). $$
 * Likewise, if
   $$ n_2 p_2 \ge 10 \qquad\mbox{and}\qquad n_2(1-p_2)
   \ge 10$$
   then
   $$ p_2 \sim \mbox{Norm}\left(p_2,
   \sqrt{\displaystyle{\frac{p_2(1-p_2)}{n_2}}}\right). $$

Under these conditions, the theorem tells us
$$ \widehat p_1 - \widehat p_2 \sim \mbox{Norm}\left(p_1 - p_2,
  \sqrt{\displaystyle{\frac{p_1(1-p_1)}{n_1}
  + \frac{p_2(1-p_2)}{n_2}}}\right).$$
This is a statement about the sampling distribution for
$\widehat p_1 - \widehat p_2$---that (under conditions) it is
approximately normal.  Thus, the *spread* of that sampling distribution
is rightly called the **standard error** of $\widehat p_1 -
\widehat p_2$:
$$ \mbox{SE}_{\widehat p_1 - \widehat p_2} =
  \sqrt{\displaystyle{\frac{p_1(1-p_1)}{n_1}
  + \frac{p_2(1-p_2)}{n_2}}}.$$
  
## Confidence Intervals for $p_1 - p_2$
It's going to be the usual thing:
$$ \mbox{(point estimate)} \pm (z^*)
  (\mbox{SE}_{\widehat p_1 - \widehat p_2}) $$
or, adapting to our situation (and the fact that we do not know
the values of $p_1$, $p_2$):
$$ (\widehat p_1 - \widehat p_2) \pm (z^*)
  \sqrt{\displaystyle{\frac{\widehat p_1(1-\widehat p_1)}{n_1}
  + \frac{\widehat p_2(1-\widehat p_2)}{n_2}}}.$$
  
Examples:

1. **One True Love** (see Example 6.19).  Here (summarized data)
  $$ \widehat p_f = \frac{363}{1412} \doteq 0.257
  \qquad\mbox{and}\qquad
  \widehat p_m = \frac{372}{1213} \doteq 0.307. $$

2. **Scolding Crows** (see Data 6.3).  Here (summarized data)
  $$ \widehat p_1 = \frac{158}{444} \doteq 0.356
  \qquad\mbox{and}\qquad
  \widehat p_2 = \frac{109}{922} \doteq 0.118. $$
  Group 1 represents the "taggers".

3. **KidsFeet** (available when Mosaic package is loaded).
  Here, we have raw data on variables `biggerfoot` and `domhand`.
  

\newpage
Carrying out the **Scolding Crows** example, we have

 * point estimate
```{r}
pointEst <- 158/444 - 109/922
pointEst
```
 * standard error
```{r}
se = sqrt(158/444*(1 - 158/444)/444 + 109/922*(1-109/922)/922)
se
```
 * z*-value
```{r}
zstar <- qnorm(.99)
zstar
```

And the 98% CI is
```{r}
pointEst + c(-1,1)*zstar*se
```

If we use `prop.test()` as a one-stop-shopping method to solve
(saving us from the individual calculations)
```{r}
prop.test(c(158,109), c(444,922), conf.level=.98)
```

