---
title: "Doing Numerical Integration in RStudio"
author: "T.Scofield"
format: pdf
---

```{r}
#| include: false
require(mosaic)
require(mosaicCalc)
```

# Definite integrals

For a built-in function like $\sin(x)$, integrating from $0$ to $\pi$
is as simple as
```{r}
integrate(sin, 0, pi)
```

But, there may not *be* a built-in R function that suits your purposes.
For instance, the pdf for exponential distributions is
$$ f(x) \;:=\; \left\{ \begin{array}{ll} 0, & \mbox{if}\; x < 0 \\
  \lambda e^{-\lambda x}, & \mbox{if}\; x \ge 0
  \end{array}\right. $$
implemented in R as `dexp()`.  However, the integral one must calculate
to find the mean involves the product of $x$ with this function, and
R does not have a built-in function for that.

When there is no built-in function, you have a couple of options for
defining it yourself.  One method is to use the native way R has for
defining complicated user-defined functions:
```{r}
#| eval: false
meanIntegrand = function(x, lam) {
  return( (x>=0) * x * lam*exp(-lam*x) )
}
```

The other method is to use `makeFun()`, suitable when the function
definition is less-complicated, as it is here, where I set $\lambda=0.2$:
```{r}
meanIntegrand <- makeFun( x*dexp(x, 0.2) ~ x )
```

Once defined, you can use `integrate()` to calculate definite integrals.
It is even possible to use $(-\infty)$ or $\infty$ as endpoints of
integration:
```{r}
integrate(meanIntegrand, -Inf, Inf)
```


# Indefinite integration (i.e., antidifferentiation)

The `antiD()` command is provided in the **mosaicCalc** package.
Make sure you have it loaded before trying things such as the following.

The idea is to find an antiderivative for a given function, for purposes
such as evaluating that antiderivative at various inputs, or displaying
a graph of the antiderivative.  Let's take $f(x) = \sin x$, a function
we understand.  We can produce an antiderivative, naming it $F(x)$, via

```{r}
F <- antiD(sin(x) ~ x)    # There are many, with F(x) = -cos(x) + C
```

It can be evaluated:
```{r}
F(3*pi/2) - F(0)
```

It can also be plotted:
```{r}
#| fig-width: 5
#| fig-height: 2
gf_fun( F(x) ~ x, xlim = c(-3*pi/2, 3*pi/2))
```

On p. 133, Pruim introduces the function
$$ f(x) = \left\{ \begin{array}{ll} x^2/9, & 0 \le x \le 3\\
  0, & \mbox{otherwise} \end{array} \right. $$
which serves as a pdf because it is nonnegative and the total
area under the curve is 1.  We can produce its cdf and plot
the two functions together:
```{r}
#| fig-width: 6.5
#| fig-height: 2.5
require(gridExtra)     # package provides grid.arrange() for side-by-side plotting
myf <- makeFun( (x>=0 & x<=3) * x^2/9 ~ x )
myF <- antiD( myf(x) ~ x )
p1 <- gf_fun(myf(x) ~ x, xlim = c(-1, 4)) |> gf_labs(title="pdf")
p2 <- gf_fun(myF(x) ~ x, xlim = c(-1, 4)) |> gf_labs(title="cdf")
grid.arrange(p1, p2, nrow=1)
```

