---
title: "Some Important Distributional Families"
format: pdf
geometry:
 - top=20mm
 - left=20mm
 - heightrounded
editor: visual
---

```{r}
#| include: false
require(mosaic)
require(mosaicCalc)
require(gridExtra)
require(abd)
require(triangle)
ssurv <- read.csv("https://scofield.site/teaching/data/csv/ssurv.csv")
spruce <- read.csv("https://scofield.site/teaching/data/csv/hesterberg/Spruce.csv")
wineAndDeath <- read.csv("https://scofield.site/teaching/data/csv/heartDiseaseDeathsAndWine.csv")
mlb18hitting <- read.csv("https://scofield.site/teaching/data/csv/mlb18abEligible.csv")
miaa05baskeball <- read.csv("https://scofield.site/teaching/data/csv/stob/miaa05.csv")
```

# Some Familiar(?) Ones

```{r}
#| fig-width: 7
#| fig-height: 4
#| warning: false
# triangle distributions are made available loading the triangle package
p1 <- gf_dist("unif", params = list(min = 1, max = 4), xlim = c(-1, 6),title = "Unif(1,4)")
p2 <- gf_dist("exp", rate = 4, title = "Exp(4)")
p3 <- gf_dist("norm", params = list(mean=500, sd=100), title = "Norm(500, 100)")
p4 <- gf_dist("triangle", params = list(a=0, b=4, c=1), title = "Tri(0, 4, 1)")

# grid.arrange() requires you load the gridExtra package
grid.arrange(p1,p2,p3,p4,ncol=2)
```

\newpage

For $X\sim$ Unif$(a, b)$, the pdf, mgf, mean and variance are,
respectively,
\begin{eqnarray*}
  f_X(x; a, b) &=& \frac{1}{b-a}, \qquad a \le x \le b \\[6pt]
  M_X(t) &=& \left\{\begin{array}{ll}
    \displaystyle{\frac{e^{bt}-e^{at}}{t(b-a)}}, & t\ne 0 \\
    1, & t=0 \end{array}\right. \\[6pt]
  \mbox{E}(X) &=& \frac{1}{2}(a + b) \\[6pt]
  \mbox{Var}(X) &=& \frac{1}{12}(b-a)^2
\end{eqnarray*}

For $X\sim$ Exp$(\lambda)$, the pdf, mgf, mean and variance are
\begin{eqnarray*}
  f_X(x; \alpha, \lambda) &=& \lambda e^{-\lambda x}, \qquad x > 0 \\[6pt]
  M_X(t) &=& \frac{\lambda}{\lambda-t} \\[6pt]
  \mbox{E}(X) &=& \frac{1}{\lambda} \\[6pt]
  \mbox{Var}(X) &=& \frac{1}{\lambda^2}
\end{eqnarray*}

For $X\sim$ Norm$(\mu,\sigma)$, the pdf, mgf, mean and variance are
\begin{eqnarray*}
  f_X(x; \alpha, \lambda) &=& \frac{1}{\sigma\sqrt{2\pi}}\,
    \exp\left(-\frac{1}{2}\left(\frac{x-\mu}{\sigma}\right)^2\right) \\[6pt]
  M_X(t) &=& \exp\left(\mu t + \frac{\sigma^2 t^2}{2}\right) \\[6pt]
  \mbox{E}(X) &=& \mu \\[6pt]
  \mbox{Var}(X) &=& \sigma^2
\end{eqnarray*}

For $X\sim$ Tri$(a,b,c)$, the pdf, mean and variance are
\begin{eqnarray*}
  f_X(x;a,b,c) &=& \left\{\begin{array}{ll}
    \displaystyle{\frac{2(x-a)}{(b-a)(c-a)}}, & a\le x \le c \\[10pt]
    \displaystyle{\frac{2(b-x)}{(b-a)(c-a)}}, & c < x \le b \\[10pt]
    0, & \mbox{otherwise} \end{array}\right. \\[6pt]
  \mbox{E}(X) &=& \frac{1}{3}(a + b + c) \\[6pt]
  \mbox{Var}(X) &=& \frac{1}{18}(a^2+b^2+c^2-ab-ac-bc)
\end{eqnarray*}

\newpage

# Gamma distributions

```{r}
#| fig-width: 7
#| fig-height: 4
#| warning: false
p1 <- gf_dist("gamma", params = list(shape = 2, rate = 1), title = "Gamma(2,1)")
p2 <- gf_dist("gamma", params = list(shape = 15, rate = 1), title = "Gamma(15,1)")
p3 <- gf_dist("gamma", params = list(shape = 2, scale = 10), title = "Gamma(2,10)")
p4 <- gf_dist("gamma", params = list(shape = 15, scale = 10), title = "Gamma(15,10)")
grid.arrange(p1,p2,p3,p4,ncol=2)
```

For $X\sim$ Gamma$(\alpha,\lambda)$, the pdf, mgf, mean and variance are
\begin{eqnarray*}
  f_X(x; \alpha, \lambda) &=& \frac{\lambda^\alpha}{\Gamma(\alpha)}\,x^{\alpha-1}
    e^{-\lambda x}, \qquad x > 0 \\[6pt]
  M_X(t) &=& \left(\frac{\lambda}{\lambda-t}\right)^\alpha \\[6pt]
  \mbox{E}(X) &=& \frac{\alpha}{\lambda} \\[6pt]
  \mbox{Var}(X) &=& \frac{\alpha}{\lambda^2}
\end{eqnarray*}

\newpage

# Weibull distributions

```{r}
#| fig-width: 7
#| fig-height: 4
#| warning: false
p1 <- gf_dist("weibull", params = list(shape = 2, scale = 1),title = "Weibull(2,1)")
p2 <- gf_dist("weibull", params = list(shape = 5, scale = 1),title = "Weibull(5,1)")
p3 <- gf_dist("weibull", params = list(shape = 2, scale = 10),title = "Weibull(2,10)")
p4 <- gf_dist("weibull", params = list(shape = 5, scale = 10),title = "Weibull(5,10)")
grid.arrange(p1,p2,p3,p4,ncol=2)
```

For $X\sim$ Weibull$(\alpha,\lambda)$, the pdf, mean and variance are
\begin{eqnarray*}
  f_X(x; \alpha, \beta) &=& \frac{\alpha}{\beta^\alpha}\, x^{\alpha-1}
    \exp\left(-\left(\frac{x}{\beta}\right)^\alpha\right), \qquad x > 0 \\[6pt]
  \mbox{E}(X) &=& \beta\,\Gamma\left(1 + \frac{1}{\alpha}\right) \\[6pt]
  \mbox{Var}(X) &=& \beta^2\,\left[\Gamma\left(1 + \frac{2}{\alpha}\right) -
    \left[\Gamma\left(1 + \frac{1}{\alpha}\right)\right]^2\right]
\end{eqnarray*}
\newpage

# Beta distributions

```{r}
#| fig-width: 7
#| fig-height: 4
#| warning: false
p1 <- gf_dist("beta", params = list(shape1 = 2, shape2 = 2), title = "Beta(2,2)")
p2 <- gf_dist("beta", params = list(shape1 = 2, shape2 = 0.9), title = "Beta(2,0.9)")
p3 <- gf_dist("beta", params = list(shape1 = 4, shape2 = 2), title = "Beta(4,2)")
p4 <- gf_dist("beta", params = list(shape1 = 0.9, shape2 = 0.85), title = "Beta(0.9,0.85)")
grid.arrange(p1,p2,p3,p4,ncol=2)
```

For $X\sim$ Beta$(\alpha,\beta)$, the pdf, mgf, mean and variance are
\begin{eqnarray*}
  f_X(x; \alpha, \beta) &=& \frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}
    \,x^{\alpha-1}(1-x)^{\beta - 1}, \qquad 0 \le x \le 1 \\[6pt]
  M_X(t) &=& 1 + \sum_{k=1}^\infty \left(\prod_{r=0}^{k-1} \frac{\alpha+r}
    {\alpha+\beta+r}\right)\,\frac{t^k}{k!} \\[6pt]
  \mbox{E}(X) &=& \frac{\alpha}{\alpha + \beta} \\[6pt]
  \mbox{Var}(X) &=& \frac{\alpha\beta}{(\alpha + \beta)^2(\alpha+\beta+1)}
\end{eqnarray*}

\newpage

# Binomial distributions

```{r}
#| fig-width: 7
#| fig-height: 4
#| warning: false
p1 <- gf_dist("binom", size = 10, prob = 0.5, title = "Binom(10, 0.5)")
p2 <- gf_dist("binom", size = 10, prob = 0.05, title = "Binom(10, 0.05)")
p3 <- gf_dist("binom", size = 100, prob = 0.5, title = "Binom(100, 0.5)")
p4 <- gf_dist("binom", size = 100, prob = 0.05, title = "Binom(100, 0.05)")
grid.arrange(p1,p2,p3,p4,ncol=2)
```

For $X\sim$ Binom$(n,\pi)$, the pmf, mgf, mean and variance are
\begin{eqnarray*}
  f_X(x; n, \pi) &=& \frac{n!}{x!(n-x)!} \,\pi^x (1-\pi)^{n-x},
    \qquad x = 0, 1, 2, \ldots, n \\[6pt]
  M_X(t) &=& (\pi e^t + 1 - \pi)^n \\[6pt]
  \mbox{E}(X) &=& n\pi \\[6pt]
  \mbox{Var}(X) &=& n\pi(1-\pi)
\end{eqnarray*}

\newpage

# Poisson distributions

```{r}
#| fig-width: 7
#| fig-height: 4
#| warning: false
p1 <- gf_dist("pois", lambda = 1, title = "Pois(1)")
p2 <- gf_dist("pois", lambda = 5, title = "Pois(5)")
p3 <- gf_dist("pois", lambda = 15, title = "Pois(15)")
p4 <- gf_dist("pois", lambda = 50, title = "Pois(50)")
grid.arrange(p1,p2,p3,p4,ncol=2)
```

For $X\sim$ Pois$(\lambda)$, the pmf, mgf, mean and variance are
\begin{eqnarray*}
  f_X(x; \lambda) &=& e^{-x}\,\frac{\lambda^x}{x!},
    \qquad x = 0, 1, 2, \ldots \\[6pt]
  M_X(t) &=& e^{\lambda(e^t-1)} \\[6pt]
  \mbox{E}(X) &=& \lambda \\[6pt]
  \mbox{Var}(X) &=& \lambda
\end{eqnarray*}