---
title: "Assignment: Personal Log of Commands"
author: "(your name here)"
output: pdf_document
---

```{r include=FALSE}
require(mosaic)
require(oibiostat)    # for Stat 145 students
require(Lock5withR)    # for Stat 243 students
```

# TASK: Your own personal log of commands
Give yourself examples of commands you've seen, ones which perform
various tasks.  I have included several examples discussed in
class---those for making a two-way table (between two
variables, usually categorical, in a data frame), and how to compute
means/averages.  The empty sections that follow have
headings indicating commands you are expected to know.  Give
explanations and examples as appropriate---as if **the audience is
yourself** several weeks or months from now.

The product of this work is the document itself, produced using
R Markdown, which you can refer to as needed later, and expand as you
learn more commands in the weeks to follow.  But you must also
submit a PDF built when you "Knit".  That document will be a PDF
so long as you specify pdf format in the YAML (see Line 4 in the
source .Rmd file).  But, upon rendering it, the file will be in
your file space on the server.  You'll need to download it to a
local computer, and then upload it to MOM as your homework submission.

### Make a two-way table: `tally()`

This is how I make a two-way table from the variables sex and substance found in the data frame HELPrct, supplied when you load the mosaic package.

```{r}
tally(~ substance | sex, data=HELPrct)
```

### Compute a mean/average: `mean()`

Working with the iris data set, I may want to know the mean petal
length (variable is named `Petal.Length`).  To obtain one average
for all plants in the data frame:

```{r}
mean(~ Petal.Length, data=iris)
```

If what I really want is average petal lengths broken down by
species, that becomes a bivariate issue.  I'm investigating
whether there is an association between `Species` and `Petal.Length`.
Do different `Species` seem to have different means?

```{r}
mean(~ Petal.Length | Species, data=iris)
```

### Compute a median for univariate quantitative data: `median()`

### Compute quantiles for univariate quantitative data: `qdata()`

### Construct a frequency table, univariate data: `tally()`

### Get extra information about a data set or command: `help()`

### Display several lines of a data frame: `head()`

### Display the column headings of a data frame: `names()`

### Display the data sets provided by a package: `data()`

### Display a frequency bar chart of univariate/bivariate categorical data: `gf_bar()`

### Display a relative frequency bar chart: `gf_props()`

### Display a dotplot of univariate quantitative data: `gf_dotplot()`

### Display a histogram of univariate quantitative data: `gf_histogram()`

### Compare histograms across different groups: `gf_histogram()`

### Create a container with specified contents: `c()`

### Count cases in a data frame: `nrow()`

### Select out cases from a data frame based on a selection criterion: `filter()` or `subset()`

### Read a .csv file from a specified url and storing it as a data frame: `read.csv()`

